---
title: "Wkshop1_Wrangling_Plotting"
format: html
editor: visual
---

Foundations of Data Science 
Wrangling - systematic pipeline of reading, inspecting, parsing, and shaping data

Project Initialisation and Workspace architecture 

Forking and Cloning a GitHub Repository

Forking (Connected Copy creates a copy of the repository in your GitHub account. This allows you to freely experiment with changes without affecting the original project.

Cloning (Standalone) creates a local copy of the repository on your computer. This allows you to work on the project offline and make changes locally before pushing them back to the remote repository.

```{r}
 
#usethis::use_git()

```

Workflow:
On GitHub: Copy URL > In R Studio: File > New Project > Version Control > Git > Paste URL > Name Project > Create Project
Once Project cloned, copy files to YOUR Own Project folder and commit changes to GitHub

```{r}

#remove original connection to source GitHub repository
gert::git_remote_remove("origin")

#push fresh copy to personal github 
usethis::use_github()

#update .gitignore file to ignore files and folders that should not be tracked by Git
```

Memory Allocation & Environmental Hygiene 
Disk(Permanent Storage) vs Session Environment (RAM/Computer Memory)
# purge .RData
```{r}
#House keeping
## Inventory every active object currently residing in session RAM
objects()

## Purge global environment
rm(list = ls())

## Confirm that global session memory is completely vacant 
objects()

unlink("~/.RData") #delete .RData file from disk to prevent automatic loading of objects into session memory on RStudio startup")

```

Ingesting Data into R 
CSV (Comma-Separated Values)
TSV (Tab-separated Values) = with spaces \t
Excel (Spreadsheets, Metadata formatting) 
```{r}
#Load primary data science framework and Excel import library
library(tidyverse)
library(readxl)
library(here)
```

```{r}
#Loading .CSV plain text file
benthic_cover <- read_csv(here::here("data/reef_cover_log.csv"))

#Loading .TXT/.TSV tab-separated text files
acoustic_stream <- read_tsv(here::here("data/acoustic_telemetry_stream.txt"))

#Loading .XLSX spread sheet files
fisheries_annual <- read_excel(here::here("data/fish_catch_data.xlsx"), sheet = "Commercial_2026")


```







