---
title: "Wkshop1_Wrangling_Plotting"
format: html
editor: visual
---

Foundations of Data Science Wrangling - systematic pipeline of reading, inspecting, parsing, and shaping data

Project Initialisation and Workspace architecture

Forking and Cloning a GitHub Repository

Forking (Connected Copy creates a copy of the repository in your GitHub account. This allows you to freely experiment with changes without affecting the original project.

Cloning (Standalone) creates a local copy of the repository on your computer. This allows you to work on the project offline and make changes locally before pushing them back to the remote repository.

```{r}
 
#usethis::use_git()

```

Workflow: On GitHub: Copy URL \> In R Studio: File \> New Project \> Version Control \> Git \> Paste URL \> Name Project \> Create Project Once Project cloned, copy files to YOUR Own Project folder and commit changes to GitHub

```{r}

#remove original connection to source GitHub repository
#gert::git_remote_remove("origin")

#push fresh copy to personal github 
#usethis::use_github()

#update .gitignore file to ignore files and folders that should not be tracked by Git
```

Memory Allocation & Environmental Hygiene Disk(Permanent Storage) vs Session Environment (RAM/Computer Memory) \# purge .RData

```{r}
#House keeping
## Inventory every active object currently residing in session RAM
objects()

## Purge global environment
rm(list = ls())

## Confirm that global session memory is completely vacant 
objects()

unlink("~/.RData") #delete .RData file from disk to prevent automatic loading of objects into session memory on RStudio startup")

```

Ingesting Data into R CSV (Comma-Separated Values) TSV (Tab-separated Values) = with spaces \t
Excel (Spreadsheets, Metadata formatting)

```{r}
#Load primary data science framework and Excel import library
library(tidyverse)
library(readxl)
library(here)
```

```{r}
#Loading .CSV plain text file
benthic_cover <- read_csv(here::here("data/workshop1/reef_cover_log.csv"))

#Loading .TXT/.TSV tab-separated text files
acoustic_stream <- read_tsv(here::here("data/workshop1/acoustic_telemetry_stream.txt"))

#Loading .XLSX spread sheet files
fisheries_annual <- read_excel(here::here("data/workshop1/fish_catch_data.xlsx"), sheet = "Commercial_2026")

```

Data cleaning: Loading excel data sheets with comments/field notes at top: end up as column headers

```{r}

#Read in mangrove_data

mangrove_data <- read_csv(file = here::here("data/workshop1/mangrove_survey_raw.csv"))

#Warning pops up
```

```{r}
# Use args within read_csv to skip headers and declare missing flags

mangrove_data <- read_csv(here::here("data/workshop1/mangrove_survey_raw.csv"), 
                           skip = 5, #skip first 5 lines of field notes
                           na = c(".", "NA", "9999", "ND", "blank")) #convert known text alts to true NA
mangrove_data #fixed data: now shows appropriate column headers and missing values
             
```

Data Frame Architecture: Tibbles(tidyverse not base R) vs Legacy(Base R) Tables Functional Behaviour: #Console Output Legacy Base R data.frame - print EVERY column & row Modern Tidyverse tibble - print only first 10 rows and columns that fit on screen

#Vector Verification Base R - print only text names(columns), requires different function to view variables e.g., str() or class() Tidy tibble - print column names and variable vector type e.g., <chr>, <dbl>, <fct>

#Partial Matching Trap Base R - does not require EXACT column name, allows salinity_ppt = df\$sal, i.e., errors Tidy tibble - requires exact column name to be used

```{r}
#Force a modern tibble to degrade into a legacy base R dataframe structure 

benthic_cover_df <- as.data.frame(benthic_cover)
 

#Print old-style dataframe structure
print(benthic_cover_df) #print shows all columns and rows, no vector type info, allows partial matching

#Tibble alternative
print(benthic_cover) #print shows vector types

                           
```

Wrangling Out Ecological Signals Using Palmer Penguins (Multi-species, morphological data, Antarctica)

```{r}
#Install data package if not done already 
# install.packages("palmerpenguins")

#Load dataset into active memory
library(palmerpenguins)
data("penguins")

#Examine structure of dataset 
glimpse(penguins) #tidyverse version from dplyr package
str(penguins) #base R
summary(penguins) #generate an exploratory summary matrix

```

Foundational Grammar: Slicing, Filtering, Sorting & Transforming dplyr package : easy data manipulation input dataframe as primary argument \> outputs modified tibble

```{r}
#Vertical slice specific morphometric variables by explicit name
morphology_metric <- select(penguins, 
                            species, 
                            bill_length_mm, 
                            bill_depth_mm, 
                            body_mass_g)

glimpse(morphology_metric)
```

Isolating Attribute with function: select() \*targets variables vertically

```{r}
#Retain a continuous block of attributes using the colon operator
spatial_block <- select(penguins, species:island)
spatial_block #only top 10 rows printed

#Discard logistics tracking attributes while preserving everything else
#Use minus sign 
clean_scientific_fields <- select(penguins, -year) #specify which "columns" to remove
clean_scientific_fields

```

Sifting rows with function: filter() \*targets variables horizontally

```{r}
#Isolate observations belonging to a single categorical target group 
adelie_cohort <- filter(penguins, species == "Adelie")
adelie_cohort

#Sift out individuals using continuous numerical boundary thresholds (ranges)
#Preserve only large penguins with mass >4500g
heavy_penguins <- filter(penguins, body_mass_g >4500)
heavy_penguins

#Combine multiple conditional parameters across separate attributes
#Preserve records matching Gentoo penguins samples explicitly on Biscoe Island
biscoe_gentoo <- filter(penguins, species == "Gentoo", island == "Biscoe")
biscoe_gentoo

#Sift records matching multiple targeting flags within an explicit set
#Filtering for multiple levels in a variable
sub_islands <- filter(penguins, island %in% c("Dream", "Torgersen"))
sub_islands

```

Ordering Sequences with function: arrange() \*alter sorting configuration ofg rows within dataframe without changing individual cell values #useful for assessing size hierarchies or chronological ordering long-term environmental observations

```{r}
#Sort penguins by ascending body mass (Smallest first * default)
lightest_first <- arrange(penguins, body_mass_g)
lightest_first

#Sort penguins in descending sequence
#Use dec() layout wrapper
heaviest_first <- arrange(penguins, desc(body_mass_g))
heaviest_first

#Execute nested sorting criteria: Group by species, then sort by descending bill length
stratified_morphology <- arrange(penguins, species, desc(bill_length_mm))
stratified_morphology
```

Introducing the Pipe \|\> (Base R) or %\>% (Tidyverse, magrittr package)

#Logic of the Pipe \|\> - Without a pipe: read code "inside-out". data in middle \> mutate() \> wrap using filter() \> final wrap using select()

- With a pipe: read code from left-right or top-bottom. data \> filter \> mutate to create new column \> select columns needed

Instead of:

```{r}

penguins_subset <- mutate(penguins, bill_ratio = bill_length_mm / bill_depth_mm)
penguins_final <- filter(penguins_subset, species == "Adelie")

```

Write:

```{r}

penguins_final <- penguins |>
  mutate(bill_ration = bill_length_mm / bill_depth_mm) |>
  filter(species == "Adelie")

```

Computing New Attributes with function: mutate() \* modify existing attribute or appemd entirely new vectors to dataframe Calculating new variables based on existing raw data E.g., standard geometric conversions or compute morphological ratios

```{r}

# Calculate a new morphological ratio in our environment
penguin_ratios <- penguins  |> 
  mutate(body_mass_kg = body_mass_g / 1000,   # Convert grams to kilograms
         bill_ratio = bill_length_mm / bill_depth_mm  # Bill ratio
  )

# View your newly engineered variables appended to the far-right columns
glimpse(penguin_ratios)

```

Data Aggregation and Ecological Summarisation Isolating and sorting individual rows / columns = data cleaning Extracting meaningful biological data e.g., morphological allometry varies by species or island group = compress observations into population summary metrics

#Dual-layered combination of functions: group_by() - does not alter appearance of table summarise() - calculate targeted statistical summary tibble

```{r}
# Grouping our active memory penguins by species
grouped_penguins <- group_by(penguins, species)
grouped_penguins

# Notice that the table looks identical, but metadata notes 'Groups: species [3]'

# Collapsing the buckets into explicit summary metrics
species_mass_summary <- summarise(grouped_penguins,
  mean_mass_g = mean(body_mass_g)
)
species_mass_summary #NA - missing value trap(MVT) * a single NA disrupt mathematical aggregation functions e.g., mean(), sd(), sum(), only returning NA

#Override NA block = explicitly declare that missing cells should be dropped during calculation 
#using magrittr pipe |> , to wrangle data, avoiding MVT

#Overcoming MVT using function na.rm = TRUE (NA removal)
biological_signal <- penguins %>%
  group_by(species, sex) %>% 
  summarise(
    sample_size = n(), # Count total individuals per category 
    mean_mass_g = mean(body_mass_g, na.rm = TRUE), #ignoring missing cells
    sd_mass_g = sd(body_mass_g, na.rm = TRUE) #SD calculation
  )
biological_signal #clean overview of sexual dimorphism and species variations across the study area


```

Integrating Data Grammar with Visual Diagnostics in .qmd

- Data Layer - source dataset table object e.g., ggplot(data = ..)
- Aesthetic Mapping - define variables mapped to axes e.g., colours, shapes, size etc.
- Geometric layer - define visual shape values represent e.g, points, bars, lines, boxplots

Wrangling and Plotting in Parallel

Challenge: Visualising Body Size and Shape - Summary Table - Visualisation

```{r}

data <- penguins %>%
  group_by(species, island) %>% 
  summarise(
    sample_size = n(), # Count total individuals per category 
    mean_mass_g = mean(body_mass_g, na.rm = TRUE), #ignoring missing cells
    sd_mass_g = sd(body_mass_g, na.rm = TRUE) #SD calculation
  )
data 

ggplot(data = penguins) +   
  geom_boxplot(mapping = aes(x = species, y = body_mass_g, fill = island))

```

Piping Directly to Visualisation \* efficient workflow

Visualising Mean Trends with Uncertainty Pipe \|\> or %\>% grouped data directly into a plot and use geom_errorbar() to represent spread of data

```{r}
#Pipe directly from aggregation to plotting with error bars

mass_compare_plot <- penguins |> 
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE), 
    sd_mass = sd(body_mass_g, na.rm = TRUE), 
    n = n(),
    .groups = "drop"
  ) |>
ggplot(aes(x = species, y = mean_mass, colour = island))+
  geom_point(size = 2)+
  geom_errorbar(aes(ymin = mean_mass - sd_mass, 
                    ymax = mean_mass + sd_mass), 
                width = 0.2)+
  labs(title = "Mean Body Mass by Species and Island",
       subtitle = "Error bars represent standard deviation", 
       x = "Species",
       y = "Mean Body MAss (g)")+
  theme_minimal()

mass_compare_plot

```

Challenge: In your .qmd template, try creating a new plot. Experiment with swapping sd_mass (standard deviation) for standard error—you will need to divide your sd by the square root of n (sd / sqrt(n)). See how that changes your visual interpretation of the data uncertainty. Present the summary table alongside the associated figure and write down your thoughts on the underlying ecological story that the data are telling.

Saving, Exporting and Version Milestones

```{r}
#Create output directories if they do not exist
#if (!dir.exists("outputs/figures")) dir.create("outputs/figures") # folder for figs
#if (!dir.exists("outputs/tables")) dir.create("outputs/tables") # folder for tables
#if (!dir.exists("Rdata")) dir.create("Rdata") # folder for Rdata objects

```

TABLES - Universal CSV text file : collaborating outside of R - Native R Binary RDS object: preserves vector configuration for R

```{r}
# 1. Exporting our collapsed summary table as a universal flat text file
write_csv(biological_signal, "Mod2_R4MarineScience/output/penguin_species_mass_summary.csv")

# 2. Saving as a native R binary file
saveRDS(biological_signal, "Mod2_R4MarineScience/output/penguin_species_mass_summary.rds")


```

FIGURES - ggplot2 function: ggsave()

```{r}
ggsave("Mod2_R4MarineScience/output/mass_compare_plot.png", 
       plot = mass_compare_plot, 
       width = 120, 
       height = 120, 
       units = "mm")
```
