---
title: "Wkshop1_Wrangling_Plotting"
format: html
editor: visual
---

Foundations of Data Science 
Wrangling - systematic pipeline of reading, inspecting, parsing, and shaping data

Project Initialisation and Workspace architecture 

Forking and Cloning a GitHub Repository

Forking (Connected Copy creates a copy of the repository in your GitHub account. This allows you to freely experiment with changes without affecting the original project.

Cloning (Standalone) creates a local copy of the repository on your computer. This allows you to work on the project offline and make changes locally before pushing them back to the remote repository.

```{r}
 
#usethis::use_git()

```

Workflow:
On GitHub: Copy URL > In R Studio: File > New Project > Version Control > Git > Paste URL > Name Project > Create Project
Once Project cloned, copy files to YOUR Own Project folder and commit changes to GitHub

```{r}

#remove original connection to source GitHub repository
gert::git_remote_remove("origin")

#push fresh copy to personal github 
usethis::use_github()

#update .gitignore file to ignore files and folders that should not be tracked by Git
```

Memory Allocation & Environmental Hygiene 
Disk(Permanent Storage) vs Session Environment (RAM/Computer Memory)
# purge .RData
```{r}
#House keeping
## Inventory every active object currently residing in session RAM
objects()

## Purge global environment
rm(list = ls())

## Confirm that global session memory is completely vacant 
objects()

unlink("~/.RData") #delete .RData file from disk to prevent automatic loading of objects into session memory on RStudio startup")

```

Ingesting Data into R 
CSV (Comma-Separated Values)
TSV (Tab-separated Values) = with spaces \t
Excel (Spreadsheets, Metadata formatting) 
```{r}
#Load primary data science framework and Excel import library
library(tidyverse)
library(readxl)
library(here)
```

```{r}
#Loading .CSV plain text file
benthic_cover <- read_csv(here::here("data/workshop1/reef_cover_log.csv"))

#Loading .TXT/.TSV tab-separated text files
acoustic_stream <- read_tsv(here::here("data/workshop1/acoustic_telemetry_stream.txt"))

#Loading .XLSX spread sheet files
fisheries_annual <- read_excel(here::here("data/workshop1/fish_catch_data.xlsx"), sheet = "Commercial_2026")

```

Data cleaning: Loading excel data sheets with comments/field notes at top: end up as column headers
```{r}

#Read in mangrove_data

mangrove_data <- read_csv(file = here::here("data/workshop1/mangrove_survey_raw.csv"))

#Warning pops up
```

```{r}
# Use args within read_csv to skip headers and declare missing flags

mangrove_data <- read_csv(here::here("data/workshop1/mangrove_survey_raw.csv"), 
                           skip = 5, #skip first 5 lines of field notes
                           na = c(".", "NA", "9999", "ND", "blank")) #convert known text alts to true NA
mangrove_data #fixed data: now shows appropriate column headers and missing values
             
```

Data Frame Architecture: Tibbles(tidyverse not base R) vs Legacy(Base R) Tables
Functional Behaviour:
#Console Output
Legacy Base R data.frame - print EVERY column & row
Modern Tidyverse tibble - print only first 10 rows and columns that fit on screen

#Vector Verification 
Base R - print only text names(columns), requires different function to view variables e.g., str() or class()
Tidy tibble - print column names and variable vector type e.g., <chr>, <dbl>, <fct>

#Partial Matching Trap
Base R - does not require EXACT column name, allows salinity_ppt = df$sal, i.e., errors
Tidy tibble - requires exact column name to be used

```{r}
#Force a modern tibble to degrade into a legacy base R dataframe structure 

benthic_cover_df <- as.data.frame(benthic_cover)
 

#Print old-style dataframe structure
print(benthic_cover_df) #print shows all columns and rows, no vector type info, allows partial matching

#Tibble alternative
print(benthic_cover) #print shows vector types

                           
```

Wrangling Out Ecological Signals Using Palmer Penguins 
(Multi-species, morphological data, Antarctica)
```{r}
#Install data package if not done already 
# install.packages("palmerpenguins")

#Load dataset into active memory
library(palmerpenguins)
data("penguins")

#Examine structure of dataset 
glimpse(penguins) #tidyverse version from dplyr package
str(penguins) #base R
summary(penguins) #generate an exploratory summary matrix

```

Foundational Grammar: Slicing, Filtering, Sorting & Transforming
```{r}


```


Isolating Attribute with function: select()
```{r}

```









