---
title: "Wkshop_EstuaryFishSurvey"
format: html
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

Estuary Fish Survey - Data Rescue Mission The Premise: You have inherited a project from a former PhD student who was studying the \*relationship between water quality and predatory fish assemblages along the Ross River Estuary gradient. Unfortunately, their data management was entirely manual, and the files are a disaster. They left you with four separate datasets: - estuary_catch_log.xlsx: Handwritten field counts of fish species (using common names), spread across multiple spreadsheet tabs. - estuary_metadata.csv: The spatial coordinates and estuary zones for each site. - estuary_sonde_data.csv: High-frequency water quality sensor readings. - species_dictionary.csv: A taxonomic lookup table linking local common names to their accepted scientific names.

Task: Write a single, reproducible R script (or .qmd) that ingests this mess, cleans it, joins it into a single master dataset, and produces a publication-ready visualisation of the estuary's ecology.

# Introduction

The aim of this analysis is to rescue and combine four messy datasets from a study of the relationship between water quality and predatory fish assemblages along the Ross River Estuary gradient.

The workflow uses mainly **tidyverse** functions to:

1. import and clean the fish catch data;
2. standardise site and species names;
3. join fish observations to the site metadata and taxonomic dictionary;
4. clean and summarise the high-frequency sonde data;
5. join water quality with fish observations; and
6. produce a publication-ready visualisation.

The original data are not modified.

# 1. Load packages

```{r}
library(tidyverse)
library(readxl)
library(here)


# Create folders for final outputs
#dir.create("output", showWarnings = FALSE)
#dir.create("figures", showWarnings = FALSE)
```

# 2. Import catch data

The fish catch data are stored across four Excel worksheets. The worksheets are combined into one data frame using `map_dfr()`.

```{r}
catch_file <- here::here("data/workshop2/estuary_catch_log.xlsx")

sheet_names <- excel_sheets(catch_file) #path must be a string

fish <- map_dfr(
  sheet_names,
  ~ read_excel(catch_file, sheet = .x)
)

fish
```

## Clean the fish data

The site and species names have inconsistent capitalisation, spaces and underscores. These are standardised before joining the datasets.

```{r}
fish <- fish |>
  mutate(
    site = str_to_lower(site),   #standardise to lowercase letters
    site = str_replace_all(site, " ", "_"), #remove _
    site = str_squish(site), #remove white spaces

    species = str_to_lower(species),
    species = str_squish(species),

    date = as.Date(date),     #set class
    count = as.numeric(count) 
  ) |>
  select(site, date, species, count) |>
  arrange(site, date, species)

fish
```

Check for missing (NAs) observations:
```{r}
fish |>
  filter(is.na(site) | is.na(date) | is.na(species) | is.na(count))
```

Remove incomplete fish observations:
```{r}
fish <- fish |>
  drop_na(site, date, species, count)
fish
```

# 3. Import and clean the species dictionary

The dictionary provides the accepted scientific name for each local common name.

```{r}
species_dictionary <- read_csv(here::here("data/workshop2/species_dictionary.csv"),
                               show_col_types = FALSE)

species_dictionary
```

Standardise the common names so they match the fish dataset.

```{r}
species_dictionary <- species_dictionary |>
  mutate(
    common_name = str_to_lower(common_name),
    common_name = str_squish(common_name)
  ) |>
  select(common_name, scientific_name) |>
  arrange(common_name)

species_dictionary
```

## Join common names to scientific names

A `left_join()` keeps every fish observation and adds the corresponding scientific name.

```{r}
fish <- fish |>
  left_join(
    species_dictionary,
    by = c("species" = "common_name")
  ) |>
  arrange(date, site, scientific_name)

fish
```

Check whether any common names failed to match:

```{r}
fish |>
  filter(is.na(scientific_name)) |>
  select(site, date, species, count)
```

# 4. Import and clean site metadata

```{r}
metadata <- read_csv(here::here("data/workshop2/estuary_metadata.csv"),
                     show_col_types = FALSE)

metadata
```

The site names use different combinations of spaces, underscores and capitalisation. First convert them to a consistent format.

```{r}
metadata <- metadata |>
  mutate(
    site_name = str_to_lower(site_name),
    site_name = str_replace_all(site_name, " ", "_")
  )
```

The site name can also be separated into two components and then recombined into a consistent `site_key`. This demonstrates `separate()` and `unite()` while making the standardisation explicit.

```{r}
metadata <- metadata |>
  separate(
    site_name,
    into = c("site_part1", "site_part2"),
    sep = "_",
    extra = "merge",
    fill = "right"
  ) |>
  mutate(
    site_part1 = str_squish(site_part1),
    site_part2 = str_squish(site_part2)
  ) |>
  unite(
    site_key,
    site_part1,
    site_part2,
    sep = "_",
    remove = FALSE
  ) |>
  select(
    site_key,
    lat,
    lon,
    zone
  ) |>
  arrange(lat)

metadata
```

# 5. Join fish observations to the site metadata

The metadata currently uses `site_key`, so create the same key in the fish data:

```{r}
fish <- fish |>
  mutate(
    site_key = site
  ) |>
  select(
    site_key,
    date,
    species,
    scientific_name,
    count
  )
```

Join the spatial information:

```{r}
fish <- fish |>
  left_join(
    metadata,
    by = "site_key"
  ) |>
  select(
    site_key,
    date,
    species,
    scientific_name,
    count,
    lat,
    lon,
    zone
  ) |>
  arrange(date, zone, scientific_name)

fish
```

Check for unmatched sites:
```{r}
fish |>
  filter(is.na(zone))
```

# 6. Import the sonde data

The sonde contains high-frequency observations collected every 15 minutes.

```{r}
sonde <- read_csv(here::here("data/workshop2/estuary_sonde_data.csv"),
                  show_col_types = FALSE)

sonde
```

## Clean the sonde data

The timestamp is converted from day/month/year format using `dmy_hm()`.

```{r}
sonde <- sonde |>
  mutate(
    site = str_to_lower(site),
    site = str_replace_all(site, " ", "_"),

    timestamp = dmy_hm(timestamp),

    temperature = as.numeric(temperature),
    salinity = as.numeric(salinity),
    turbidity = as.numeric(turbidity)
  ) |>
  select(
    site,
    timestamp,
    temperature,
    salinity,
    turbidity
  ) |>
  arrange(site, timestamp)

sonde
```

Create a date column for joining with the daily fish survey data:

```{r}
sonde <- sonde |>
  mutate(
    date = as.Date(timestamp)
  )
```

# 7. Reshape the sonde data with `pivot_longer()`

Putting the three water-quality variables into one column makes it easier to apply the same quality-control rules to all variables.

```{r}
sonde_long <- sonde |>
  pivot_longer(
    cols = c(temperature, salinity, turbidity),
    names_to = "variable",
    values_to = "value"
  ) |>
  arrange(site, timestamp, variable)

sonde_long
```

## Remove invalid sensor values

The turbidity data contain negative values including the `-999` missing-value code. Negative salinity values are also invalid.

```{r}
sonde_long <- sonde_long |>
  mutate(
    value = case_when(
      variable == "turbidity" & value < 0 ~ NA_real_,
      variable == "salinity" & value < 0 ~ NA_real_,
      variable == "temperature" & (value < 0 | value > 40) ~ NA_real_,
      TRUE ~ value
    )
  )
```

Remove observations with missing values:

```{r}
sonde_long <- sonde_long |>
  drop_na(value)
```

# 8. Calculate daily water-quality values

The fish data are recorded by day, so the 15-minute sonde observations are summarised to daily means.

```{r}
sonde_daily_long <- sonde_long |>
  group_by(site, date, variable) |>
  summarise(
    daily_mean = mean(value),
    .groups = "drop"
  ) |>
  arrange(date, site, variable)

sonde_daily_long
```

## Convert the water-quality variables back to columns

`pivot_wider()` converts the long data back into one row per site and date.

```{r}
sonde_daily <- sonde_daily_long |>
  pivot_wider(
    names_from = variable,
    values_from = daily_mean
  ) |>
  arrange(date, site)

sonde_daily
```

Rename the site column so it can be joined with the fish dataset:

```{r}
sonde_daily <- sonde_daily |>
  rename(
    site_key = site
  )
```

# 9. Join fish and water-quality data

The fish observations and daily sonde data have two common variables: `site_key` and `date`.

```{r}
master <- fish |>
  left_join(
    sonde_daily,
    by = c("site_key", "date")
  ) |>
  arrange(date, zone, site_key, scientific_name)

master
```

The resulting `master` dataset contains:

- fish abundance;
- common and scientific species names;
- site coordinates;
- estuary zone;
- daily temperature;
- daily salinity; and
- daily turbidity.

# 10. Check the rescued master dataset

Check the dimensions:

```{r}
dim(master)
```

Check the sites:

```{r}
master |>
  select(site_key, zone, lat, lon) |>
  distinct() |>
  arrange(lat)
```

Check the species:

```{r}
master |>
  select(species, scientific_name) |>
  distinct() |>
  arrange(scientific_name)
```

Check for missing joined values:

```{r}
master |>
  filter(
    is.na(scientific_name) |
    is.na(zone) |
    is.na(salinity) |
    is.na(turbidity)
  )
```

Export the rescued dataset:

```{r}

write_csv(
  master,
  here::here("output/estuary_master_dataset.csv")
)

```

# 11. Prepare fish assemblage data

Calculate the total abundance of each species within each estuary zone.

```{r}
fish_zone <- master |>
  group_by(zone, scientific_name) |>
  summarise(
    abundance = sum(count),
    .groups = "drop"
  ) |>
  arrange(zone, desc(abundance))

fish_zone
```

Calculate relative abundance within each zone:

```{r}
fish_zone <- fish_zone |>
  group_by(zone) |>
  mutate(
    relative_abundance = abundance / sum(abundance)
  ) |>
  ungroup()

fish_zone
```

# 12. Publication-ready visualisation

Summary Table 
```{r}

summary_table <- master %>%
  group_by(scientific_name, zone) %>%
  summarise(
    mean_count = mean(count, na.rm = TRUE),
    sd_count = sd(count, na.rm = TRUE),
    mean_salinity = mean(salinity, na.rm = TRUE),
    sd_salinity = sd(salinity, na.rm = TRUE),
    .groups = "drop"
  )

summary_table

```


```{r}
write_csv(
  summary_table,
  here::here("output/summary_table.csv")
)

```


The main figure shows how the fish assemblage changes along the Ross River estuary gradient.

```{r}
figure <- ggplot(fish_zone, 
                 aes(x = zone, y = relative_abundance, fill = scientific_name)) +
  geom_col(width = 0.7, colour = "white") +
  scale_y_continuous(
    labels = scales::percent,
    expand = c(0, 0)) +
  labs(
    title = "Predatory fish assemblages across the Ross River Estuary",
    subtitle = "Relative abundance of recorded fish taxa across the estuary gradient",
    x = "Estuary zone",
    y = "Relative abundance",
    fill = "Fish taxon") +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 10),
    legend.position = "bottom",
    axis.text.x = element_text(angle = 20, hjust = 1)
  )

figure
```
ePortfolio figure 
```{r}

ggplot(master,
  aes(x = salinity, y = count)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ scientific_name, scales = "free_y") +
  labs(
    title = "Fish Abundance Along the Salinity Gradient",
    x = "Salinity",
    y = "Fish abundance"
  ) +
  theme_classic() +
  theme(
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )

```

```{r}

ggplot(master, aes(x = salinity, y = count, colour = scientific_name)) +
  geom_point(size = 1.5, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  facet_wrap(
    ~ scientific_name,
    scales = "free_y") +
  scale_colour_brewer(
    palette = "Dark2") +
  labs(
    title = "Fish Abundance Along the Salinity Gradient",
    x = "Salinity",
    y = "Fish abundance") +
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold", size = 12),
    strip.text = element_text(face = "bold", size = 11),
    legend.position = "none"
  )

```



Save the figure:

```{r}
ggsave(
  here::here("output/ross_river_fish_assemblage.png"),
  figure,
  width = 10,
  height = 6,
  dpi = 300
)
```

# 13. Optional environmental relationship

A second visualisation can examine the relationship between fish abundance and salinity.

```{r}
ggplot(
  master,
  aes(x = salinity, y = count, colour = zone)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Fish abundance in relation to salinity",
    x = "Daily mean salinity",
    y = "Fish abundance",
    colour = "Estuary zone") +
  theme_classic(base_size = 12)

```

# Conclusion

The rescued dataset combines the original fish observations, taxonomic information, spatial metadata and daily water-quality measurements into a single reproducible dataset.

The main data-management steps were:

- standardising inconsistent site and species names with `mutate()`;
- selecting and arranging relevant variables with `select()` and `arrange()`;
- removing incomplete records with `drop_na()`;
- separating and recombining site identifiers with `separate()` and `unite()`;
- reshaping high-frequency water-quality measurements with `pivot_longer()` and `pivot_wider()`;
- combining datasets with `left_join()`; and
- summarising fish abundance and water quality with `group_by()` and `summarise()`.

The final master dataset is saved as `output/estuary_master_dataset.csv`, and the publication-ready fish assemblage figure is saved in the `figures/` folder.


