---
title: "Wkshop_EstuaryFishSurvey"
format: html
editor: visual
---
Estuary Fish Survey - Data Rescue Mission
The Premise: You have inherited a project from a former PhD student who was studying the 
*relationship between water quality and predatory fish assemblages along the Ross River Estuary gradient. Unfortunately, their data management was entirely manual, and the files are a disaster.
They left you with four separate datasets:
estuary_catch_log.xlsx: Handwritten field counts of fish species (using common names), spread across multiple spreadsheet tabs.
estuary_metadata.csv: The spatial coordinates and estuary zones for each site.
estuary_sonde_data.csv: High-frequency water quality sensor readings.
species_dictionary.csv: A taxonomic lookup table linking local common names to their accepted scientific names.
Your mission is to write a single, reproducible R script (or .qmd) that ingests this mess, cleans it, joins it into a single master dataset, and produces a publication-ready visualisation of the estuary's ecology.

```{r}
#load packages
library(tidyverse)
library(readxl)
library(here)

#loading datasets
# .xlsx file  estuary_catch_log = multiple sheets
lower_ross <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = "Lower_ross")
mid_ross <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = "Mid_Ross ")
ross_mouth <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = "ROSS MOUTH")
upper_ross <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = "Upper Ross")

# .csv files  (here::here) to specifiy working directory 
metadata <- read_csv(here::here("data/workshop2/estuary_metadata.csv"))
sonde <- read_csv(here::here("data/workshop2/estuary_sonde_data.csv"))
species <- read_csv(here::here("data/workshop2/species_dictionary.csv"))
```

```{r}
usethis::use_github()

```




