---
title: "Wrkshop2_Advanced_Wrangling"
format: html
editor: visual
---

Connect to github

```{r}
#usethis::use_git 

#usethis::use_github()
```

ggplot - geom_smooth method = - "lm" linear model i.e. best fit line, - "gam" generalized additive model i.e., flexible curve, se =\
- TRUE, confidence interval, - FALSE no confidence interval, formula = y \~ x, formula = y \~ poly(x, 2) quadratic model, formula = y \~ poly(x, 3) cubic model

Saving Images using ggsave

```{r}

# | fig-cap: "Title of the figure" 
# | fig-width:: number
# | fig-height:: number (both show exact size of figure in image)
# OR in {r, fig.width=6, fig.height=4}

#ggsave(plotname, filename = here::here("figures/plotname.png"), width = 6, height = 4, units = "in", dpi = 300) 
#gives image of plot created

```

Load packages - tidy formats: i.e more raw information, not rates or types: 2 of the same per variable e.g., type: #cases and population per country = 2 countries per type

```{r}
library(tidyverse)

```

Table Structure Examples

```{r}

table1 
table2
table3

```

Interrelated rules: - Each variable must have it's own column - Each observation must have its own row - Each value must have its own cell \*Consistent data structure easier for data transformations and statistical analysis

e.g.,

```{r}

# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)


# Compute cases per year
table1 %>% 
  count(year, wt = cases) 
#> # A tibble: 2 × 2
#>    year      n
#>   <int>  <int>
#> 1  1999 250740
#> 2  2000 296920

#(Frequency weight) wt = NULL if counts number of rows in each group OR 
#wt = variable is computes sum(wt) for each group

# Visualise changes over time
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))

```

Pivoting Data/ Cleaning Step 1. - Understand dataset: what each variable and observation means/ represents Step 2. - Resolving common problems: A variable is spread across multiple columns\
An observation is scattered across multiple rows i.e., pivot data / move it around

Lengthening Datasets using function: tidyr:pivot_longer() which \>rows & \<columns

```{r}

billboard #messy data: each week of the year as a column - change to 1 column for every week and 1 column for every rank

billboard |> pivot_longer(cols = starts_with("wk"),  #columns begin with "name", same as select()
                         names_to = "week",          #names column storing all "name" columns
                         values_to = "rank")         # values stored in cells of "name" column now named "ranks"

```

NAs for values wherever blanks present (may not be actual unknowns) can be removed using function: values_drop_na = TRUE)

```{r}

billboard |> pivot_longer(cols = starts_with("wk"), 
                           names_to = "week",
                           values_to = "rank",
                           values_drop_na = TRUE)
#All NAs removed & fewer rows
```

Additional Tidying \*converting strings to numbers using function: mutate() and parse_number()

```{r}

billboard_longer <- billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  ) |> 
  mutate(
    week = parse_number(week) #extracts first number from a string i.e. removed the 'wk' from wk1
  )
billboard_longer

```

Pivoting Longer e.g.,

```{r}
df <- tribble(
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)

#tidy dataset w/ 3 variable: id, measurement and value
df_tidy <- df |> pivot_longer(cols = bp1:bp2, 
                              names_to = "measurement", 
                              values_to = "value")
df_tidy #repeated id for each measurement 

```

Additional wrangling of pivot_long (many variable in column names)

```{r}
who2 |> 
  pivot_longer(
    cols = !(country:year),
    names_to = c("diagnosis", "gender", "age"), #renames each column based on ""
    names_sep = "_", #names contain multiple values, "_" controls how the column is broken down e.g. column name "sp_m_014" = VALUES sp m 014
    values_to = "count" #values from original cells 
  )

```

Widening Datasets using function: tidyr:pivot_wider() which \>columns & \<rows

```{r}
cms_patient_experience #messy dataset: start 5 columns

cms_patient_experience|> distinct(measure_cd, measure_title) #all values for these variables

cms_patient_experience|> pivot_wider(names_from = measure_cd, values_from = prf_rate) #now 9 columns

cms_patient_experience|> pivot_wider(id_cols = starts_with("org"), #specify columns that unique identify each observation i.e. value
                                     names_from = measure_cd,
                                     values_from = prf_rate)

```

Additional wrangling of pivot_wider (many variable in column names)

```{r}

df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)

df |> 
  pivot_wider(
    names_from = measurement, #creates new column for each variable in measurement: bp1, bp2, bp3
    values_from = value #matches values for each row
  )

df |> 
  distinct(measurement) |> #output = unique values of measurement
  pull()

df |> 
  select(-measurement, -value) |> #only id column 
  distinct()

df |> 
  select(-measurement, -value) |> 
  distinct() |> 
  mutate(x = NA, y = NA, z = NA) #combine results to generate empty dataframe

```

Palmer Penguins #Load penguins dataset e.g., penguins \<- read_csv (".../data/Pengu") penguins \<- read_csv(here::here("data/penguins.csv")

```{r}
library(palmerpenguins)

#Create a long version of the penguins dataset

penguins_long <- penguins |> pivot_longer(cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g),
                                          names_to = "measurement_type",
                                          values_to = "value")
head(penguins_long)

```

```{r}

penguins_long |> 
  drop_na(value) |> 
  ggplot(aes(x = value, fill = species))+
  geom_histogram(bins = 30, alpha = 0.7, colour = "black")+
  facet_wrap(~measurement_type, scales = "free_x")+
  theme_minimal()+
  labs(
    title = "Morphometric distributions across penguin species",
    x = "Measurement value",
    y = "Frequency"
  )

```

Widening Data with function: pivot_wider() \* easier to parse a wide table than a long repetitive list

```{r}

mass_summary <- penguins |> 
  drop_na(body_mass_g) |>
  group_by(species, island) |>
  summarise(mean_mass = mean(body_mass_g))

head(mass_summary) #could be tidyr

mass_matrix <- mass_summary |>
  pivot_wider(names_from = island, 
               values_from = mean_mass)

head(mass_matrix) #clear species body mass by island

```

Separating and Uniting Data Tables using function: separate()

```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/") 
#case & pop as <chr>, sep = specific character separator

table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE) 
#case & pop as <int>

table3 %>% 
  separate(year, into = c("century", "year"), sep = 2) 
#interprets integer as position to split at i.e., sep 2 = split column 2, 
#1 = far-left, -1 = far-right or the string
```

Using function unite() to merge multiple columns

```{r}
table5 %>% 
  unite(new, century, year, sep = "") #need to add sep(do not want ant separator)

```

Wrangling Strings using tidyverse::stringr (text) - Formatting mismatches e.g., "Nelly Bay", "nelly_bay", "NELLY BAY" as same location

```{r}

library(tidyverse)

# A remarkably messy data frame of field sites
messy_sites <- tibble(
  site_id = c(" Nelly Bay", "nelly_bay", "NELLY BAY", " Geoffrey_Bay ", "geoffrey bay")
  )
print(messy_sites)

# Using stringr within mutate to standardize the text
clean_sites <- messy_sites |> 
  mutate(
    #1. Convert everything to lowercase
    site_clean = str_to_lower(site_id), 
    #2. Replace any spaces with underscores
    site_clean = str_replace_all(site_clean, pattern = " ", replacement = "_"),
    #3. Trim any leading or trailing whitespaces (invisible spaces at the ends)
    site_clean = str_trim(site_clean)
  )

print(clean_sites)

```

Wrangling Dates using tidyverse::lubridate (dates) - Formatting mismatches e.g., DD/MM/YYYY - use dmy() , YYYY-MM-DD - use ymd()

```{r}
#Parsing different data formats
date_1 <- dmy("25/12/2026")
date_2 <- ymd("2026-12-25")

#R now recognises these as identical Date objects 
date_1 == date_2 #TRUE

```

Acoustic Telemetry or CTD Profiles: timestamps, appending "\_hms" (hours, minutes, seconds)

```{r}
## A tibble of raw sensor data with a messy character timestamp
sensor_data <- tibble(
  raw_time = c("14-05-2026 08:30:00", "14-05-2026 08:45:00", "14-05-2026 09:00:00"),
  temperature = c(24.5, 24.6, 24.4)
)

# Converting character strings to true POSIXct datetime objects
sensor_clean <- sensor_data |>
  mutate(
    true_time = dmy_hms(raw_time)
  )

print(sensor_clean) #changed dmy-time to ymd-time

#extract components e.g., month(true_time) to get month integer

```

Relational Data: Joining Tables using KEY column

Tables Loaded

```{r}

# Table 1: Biological observation data
observations <- tibble(
  site_code = c("NB", "GB", "MI", "NB", "HB"),
  species = c("Trout", "Snapper", "Trout", "Cod", "Trout"),
  count = c(5, 2, 1, 3, 8)
)

# Table 2: Spatial metadata
site_metadata <- tibble(
  site_code = c("NB", "GB", "MI", "RP", "WP"),
  zone = c("Marine National Park", "Conservation Park", "Habitat Protection", "General Use", "Other Use"),
  lat = c(-19.16, -19.15, -19.14, -19.12, -19.11)
)

```

Workhorse: left_join() & right_join() Function same way but in opposite directions, - L join keeps all row from left table(i.e. observations) and matches data from right table(i.e., site_metadata) wherever key matches , if missing = NA

```{r}
#Joining metadata to our observations
joined_data <- observations |> 
  left_join(site_metadata, by = join_by(site_code)) 
#values only from sites in common and added HB from observations
print(joined_data)

```

Strict Filter: inner_join() Only keeps rows that completely match in both tables Drops observations at site not listed in metadata (i.e. only for data w/ complete spatial or environmental context)

```{r}
matched_data <- observations|> 
  inner_join(site_metadata, by = join_by(site_code)) #values only from sites in common 
glimpse(matched_data)
```

Detective Filter: anti-join() Does not add new columns - Only shows rows with no match in right table(site_metadata)

```{r}
#Which observations are missinf from our metadata dictionary?
missing_context <- observations |> 
  anti_join(site_metadata, by = join_by(site_code))
glimpe(missing_context) # "HB", "Trout", "8"
```

Handling Missing Values - NAs

Converting legacy errors with: na_if() inside mutate() e.g., data from CTD casts or light loggers placeholders (-99 or -999)

```{r}
# Raw data from an old temperature logger
logger_data <- tibble(
  depth_m = c(10, 20, 30, 40),
  temp_c = c(24.5, 24.1, -999, 23.5)) # -999 is a known sensor error code

#Convert -999 error codes to true NA values
fixed_logger <- logger_data |> 
  mutate(temp_c = na_if(temp_c, -999))

print(fixed_logger)
```

Replacing Missing Values with: coalesce() inside mutate() - When NA means 0

```{r}
# Simulate some count data
shark_count <- tibble(
  site = c("Reef_A", "Reef_B", "Reef_C"),
  shark_count = c(3, NA, 5)) # The NA here actually means 0 sharks were seen

#Replace NA with 0
shark_fixed <- shark_count |>
  mutate(shark_count = coalesce(shark_count, 0))

print(shark_fixed)

#for 2 columns e.g. primary_sensor and backup_sensor, coalesce(primary_sensor, backup_sensor) will use primary reading & fill in gaps with backup reading)
```

Na Not a Number (NaN) i.e. mathematically impossible e.g. calculating Catch per unit effort (catch and effort both = 0)

```{r}

cpue_data <- tibble(
  site = c("Bay_1", "Bay_2"),
  catch = c(10, 0),
  effort_hours = c(2, 0))

# Calculating CPUE (catch / effort)
cpue_calc <- cpue_data |>
  mutate(
    cpue = catch / effort_hours)

print(cpue_calc) #can remove using drop_na() or isolate using is.nan()

```

Implicit Missing Values & Zero-data Framework: complete() e.g., knowing where a species was NOT found e i.e. presence/absence

```{r}

raw_catch <- tibble(
  site = c("Reef_1", "Reef_1", "Reef_2"),
  species = c("Pmaculatus", "Pleopardus", "Pmaculatus"),
  count = c(5, 2, 8))

# Force inclusion of hidden zero-catch data
full_catch_matrix <- raw_catch |>
  complete(site, species, fill = list(count = 0)) #fills NAs with 0 for means/averaging

print(full_catch_matrix)

```

Nuclear Option: drop_na() If NA = failed instrument reading

```{r}
sensor_log <- tibble(
  day = 1:4,
  salinity = c(35.2, 35.1, NA, 35.3)
)

# Remove any row where salinity is missing
clean_log <- sensor_log |>
  drop_na(salinity)

print(clean_log)

```

Practical Exercise: Penguins, Pivots and Joining Data
Exercise 1: Cleaning Messy Metadata
```{r}

# Make sure packages are loaded
library(tidyverse)
library(palmerpenguins)

# The messy metadata provided by your colleague
island_metadata <- tibble(
  island_name = c(" biscoe", "Dream ", "Torgersen"),
  station_install = c("15/01/2003", "22-03-2004", "05/11/2001"),
  latitude = c(-64.81, -64.73, -64.76)
)

print(island_metadata)

```
Task: Create a new object called clean_metadata. Using mutate(), stringr, and lubridate, fix the island_name column so that it perfectly matches the island column in the penguins dataset (hint: watch out for those hidden spaces!). Then, convert the station_install column into a proper Date object.
```{r}
#fixing island names
clean_metadata <- island_metadata |>
  mutate(
    island = str_to_title(island_name),
    island = str_trim(island_name),
    station_install = dmy(station_install)
  )

print(clean_metadata)

#fixing date = year 
clean_metadata |> 
  separate(station_install, into = c("year", "day", "month"), convert = TRUE) 


#joining 
joined_metadata <- penguins_long|> 
  left_join(clean_metadata, by = join_by(island)) 
#values only from sites in common and added HB from observations
print(joined_metadata)
```
