---
title: "Wrkshop4_SpatialData"
format: html
editor: visual
---

sf objsect *standards data frame with special "geometry" column attached

Installing Spatial R Packages
```{r}
  ##install packages  
#install.packages("sf") 
#install.packages("terra")
#install.packages("tmap")

#load packages
library(tidyverse)
library(sf) # simple features
library (terra) # for raster
library(tmap) # Thematic maps are geographical maps in which spatial data distributions are visualized

```

Problem: R analysis in copepod richness throughout Australia's oceans and Souther Ocean.
Hypothesis: copepode species richness will be higher in warmer waters than cooler waters. 
Develop a map that spatially displays data - is hypothesis supported/worth pursing

```{r}
#load data 
#CPR - continuous plankton recorder
copepod <- read_csv("data/data-for-course/copepods_raw.csv")
copepod

#Projection mathcing - checking coordinate refrence system (CRS) using: st_crs(sf_object) 
```

Data Exploration
```{r}
#initial plot....not a map
ggplot(copepod) + 
  aes(x = longitude, y = latitude, color = richness_raw) +
  geom_point()


#other trends/patterns 
ggplot(copepod, aes(x = latitude, y = richness_raw) + 
  stat_smooth() + 
  geom_point()

```

sf() - GIS Equivalent:shape file or feature colletion 
```{r}
sdat <- st_as_sf(copepod, coords = c("longitude", "latitude"), 
                 crs = 5126)
# st_as_sf: converst data types to simple features

# copepod: our original data

# coords: names of columns realting to spatial coordinates (X,Y)

#crs: coordinate reference system 

```

Coordinate Reference Systems 
2D mapping of 3D sphere: 
- datum: referencing point for Geographic coordinate reference systems (GCS)
- unit usually decimal degrees (longitude and latitude)
- common GCS: WGS84
- EPSG code: 5126 (Projected CRS,short hand code for specific CRS) or Well-known text (WKT) to define CRS
```{r}
crs5126 <- st_crs(5126)
crs5126 # look at the whole CRS
crs5126$Name # pull out just the name of the crs
# WGS84

```






